---
title: "Sociodemographic factors predicting healthcare utilization and insurance coverage patterns"
author: "Lovisa Johansson & Roberta Merialdo"
format: html
---

::: {.callout-note}
This report was generated using synthetic data! It is only used for demonstration purposes and does not reflect real patients or healthcare providers. The contents have not been reviewed or validated by medical professionals.
:::

## Introduction

Understanding how sociodemographic factors influence healthcare utilization and insurance coverage is important for identifying disparities in healthcare access and outcomes. Age, gender, and race are often associated with differences in health needs and healthcare-seeking behaviour. This project explores these relationships using a synthetic healthcare dataset to identify patterns that may resemble real-world inequalities in healthcare systems.

## Research Question
*"Do sociodemographic factors such as age, gender, and race predict healthcare utilization and insurance coverage patterns?"*

## Data and Methods

### Dataset Description

This analysis uses synthetic healthcare data generated by [Synthea](https://github.com/synthetichealth/synthea), a synthetic patient population simulator. The dataset contains 17 interlinked CSV files covering patients, clinical encounters, conditions, medications, procedures, and insurance information. The primary datasets used in this analysis are:

- **patients** — demographic information including age, gender, race, and insurance coverage
- **encounters** — records of all clinical visits per patient, used to measure healthcare utilization
- **payer_transitions** — records of insurance coverage changes over time

### Variable Definitions

| Variable | Source | Description |
|---|---|---|
| `age` | patients | Age in years, derived from birthdate |
| `gender` | patients | Patient gender (M/F) |
| `race` | patients | Patient race (white, black, asian, etc.) |
| `n_encounters` | encounters | Number of clinical encounters per patient (outcome 1) |
| `healthcare_coverage` | patients | Total insurance coverage amount in USD (outcome 2) |

### Statistical Methods

Two outcomes were analysed:

1. **Healthcare utilization** (`n_encounters`) — a count variable modelled using Poisson regression
2. **Insurance coverage** (`healthcare_coverage`) — a continuous variable modelled using linear regression

Prior to regression, hypothesis tests were conducted for each predictor–outcome combination (H1a–H3b):

- **Age** — Spearman correlation (non-parametric, due to skewed distributions), tested separately for each outcome
- **Gender** — Independent samples t-test, tested separately for each outcome
- **Race** — Kruskal-Wallis test (non-parametric, multiple groups), tested separately for each outcome

## Data Preparation

```{r}
#| label: data-preparation
#| message: false

library(ggplot2)
library(knitr)
library(data.table)
targets::tar_load(dt_patients)
targets::tar_load(dt_encounters)
targets::tar_load(dt_payer_transitions)

# Calculate age from birthdate
dt_patients[, age := as.numeric(difftime(Sys.Date(), as.Date(birthdate), units = "days")) / 365.25]

# Count number of encounters per patient (healthcare utilization)
encounter_counts <- dt_encounters[, .(n_encounters = .N), by = patient]

# Join patients with encounter counts and coverage
analysis <- dt_patients[encounter_counts, on = .(id = patient), nomatch = 0]
analysis <- analysis[, .(id, age, gender, race, n_encounters, healthcare_coverage)]

# Remove deceased patients
analysis <- analysis[is.na(dt_patients[match(analysis$id, dt_patients$id), deathdate]) |
                     dt_patients[match(analysis$id, dt_patients$id), deathdate] == ""]
```

### Create Analysis Dataset

### Table of Missing Values in Key Variables

```{r}
#| label: missing-values
missing <- sapply(analysis, function(x) sum(is.na(x)))
kable(data.frame(Variable = names(missing), Missing = missing), row.names = FALSE)
```

## Exploratory Data Analysis

### Sample Characteristics

```{r}
#| label: sample-characteristics
summary_table <- analysis[, .(
  N = .N,
  Mean_Age = round(mean(age, na.rm = TRUE), 1),
  SD_Age = round(sd(age, na.rm = TRUE), 1),
  Mean_Encounters = round(mean(n_encounters, na.rm = TRUE), 1),
  Mean_Coverage = round(mean(healthcare_coverage, na.rm = TRUE), 0)
)]
kable(summary_table)
```

### Visualisations

@fig-age-distribution shows the age distribution of patients. @fig-encounters-gender and @fig-encounters-race display healthcare utilization by gender and race respectively. @fig-coverage-race shows insurance coverage by race, and @fig-age-encounters illustrates the relationship between age and number of encounters.

```{r}
#| label: fig-age-distribution
#| fig-cap: "Age distribution of patients"
ggplot(analysis, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(x = "Age (years)", y = "Count", title = "Age Distribution") +
  theme_minimal()
```

```{r}
#| label: fig-encounters-gender
#| fig-cap: "Number of encounters by gender"
ggplot(analysis, aes(x = gender, y = n_encounters, fill = gender)) +
  geom_boxplot() +
  labs(x = "Gender", y = "Number of Encounters", title = "Healthcare Utilization by Gender") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#| label: fig-encounters-race
#| fig-cap: "Number of encounters by race"
ggplot(analysis, aes(x = race, y = n_encounters, fill = race)) +
  geom_boxplot() +
  labs(x = "Race", y = "Number of Encounters", title = "Healthcare Utilization by Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

```{r}
#| label: fig-coverage-race
#| fig-cap: "Insurance coverage by race"
ggplot(analysis, aes(x = race, y = healthcare_coverage, fill = race)) +
  geom_boxplot() +
  labs(x = "Race", y = "Healthcare Coverage ($)", title = "Insurance Coverage by Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

```{r}
#| label: fig-age-encounters
#| fig-cap: "Relationship between age and number of encounters"
ggplot(analysis, aes(x = age, y = n_encounters)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red") +
  labs(x = "Age (years)", y = "Number of Encounters", title = "Age vs Healthcare Utilization") +
  theme_minimal()
```

## Hypothesis Testing

### Healthcare Utilization (`n_encounters`)

#### Hypothesis 1a: Age is associated with healthcare utilization

```{r}
#| label: h1a-age-utilization
# H1a: Age is associated with healthcare utilization
cor.test(analysis$age, analysis$n_encounters, method = "spearman")
```

#### Hypothesis 2a: Gender is associated with healthcare utilization

```{r}
#| label: h2a-gender-utilization
# H2a: Gender is associated with healthcare utilization
t.test(n_encounters ~ gender, data = analysis)
```

#### Hypothesis 3a: Race is associated with healthcare utilization

```{r}
#| label: h3a-race-utilization
# H3a: Race is associated with healthcare utilization
kruskal.test(n_encounters ~ race, data = analysis)
```

### Insurance Coverage (`healthcare_coverage`)

#### Hypothesis 1b: Age is associated with insurance coverage

```{r}
#| label: h1b-age-coverage
# H1b: Age is associated with insurance coverage
cor.test(analysis$age, analysis$healthcare_coverage, method = "spearman")
```

#### Hypothesis 2b: Gender is associated with insurance coverage

```{r}
#| label: h2b-gender-coverage
# H2b: Gender is associated with insurance coverage
t.test(healthcare_coverage ~ gender, data = analysis)
```

#### Hypothesis 3b: Race is associated with insurance coverage

```{r}
#| label: h3b-race-coverage
# H3b: Race is associated with insurance coverage
kruskal.test(healthcare_coverage ~ race, data = analysis)
```

## Statistical Analysis

```{r}
#| label: regression

# Outcome 1: Healthcare utilization (count → Poisson regression)
model_utilization <- glm(n_encounters ~ age + gender + race,
                         data = analysis,
                         family = poisson())
summary(model_utilization)

# Outcome 2: Insurance coverage (continuous → linear regression)
model_coverage <- lm(healthcare_coverage ~ age + gender + race,
                     data = analysis)
summary(model_coverage)
```

## Discussion and Conclusion

### Key Findings

This analysis examined whether age, gender, and race predict healthcare utilization and insurance coverage among synthetic patients. The results suggest that age and gender are the strongest predictors of both outcomes in this dataset.

Age was positively associated with healthcare utilization. The Spearman correlation showed a strong positive relationship between age and the number of encounters, indicating that older patients tended to have more clinical visits. This finding was also supported by the Poisson regression model, in which age remained a strong and statistically significant predictor of healthcare utilization after adjustment for gender and race.

Gender was also significantly associated with healthcare utilization. Female patients had a higher mean number of encounters than male patients, and this pattern remained significant in the regression model, where male gender was associated with fewer encounters. This suggests that healthcare use differs by gender in the synthetic population.

In contrast, race was not significantly associated with healthcare utilization in the Kruskal-Wallis test. However, in the multivariable Poisson regression model, several racial categories showed statistically significant differences after adjustment for age and gender. This suggests that racial differences may become more visible when multiple sociodemographic factors are considered simultaneously.

Age was also positively associated with healthcare coverage. Older patients tended to have higher coverage levels, as shown by the Spearman correlation and confirmed by the linear regression model. Gender differences were again substantial: female patients had significantly higher average healthcare coverage than male patients, and male gender remained negatively associated with coverage in the adjusted model.

For race, no statistically significant differences in healthcare coverage were found in the Kruskal-Wallis test. In the linear regression model, most racial categories were not statistically significant at the 5% level, although the coefficient for Black patients was significant, indicating lower coverage relative to the reference category.

### Interpretation

Overall, the findings suggest that sociodemographic factors are associated with healthcare utilization and insurance coverage patterns in this synthetic dataset. In particular, age and gender appear to play an important role in shaping both healthcare use and coverage. Race showed weaker and less consistent associations, especially in the bivariate analyses, although some adjusted differences emerged in the regression models.

These results are broadly consistent with the idea that healthcare needs and access may vary across demographic groups. Older individuals may require more frequent healthcare services, while gender differences may reflect variations in healthcare-seeking behaviour, health conditions, or service use patterns.

### Limitations

- The dataset is synthetic and does not represent real patients or healthcare providers. Therefore, the findings should be interpreted as a demonstration of analytical methods rather than as evidence about real-world healthcare systems.
- The count outcome may show overdispersion, meaning that the Poisson model may not fully capture the variability in healthcare utilization.
- Healthcare coverage is highly skewed, which may affect the assumptions of linear regression.
- The analysis only included a limited number of predictors. Other relevant factors, such as socioeconomic status, comorbidities, geography, or insurance type, were not included in the final models.

### Conclusion

In conclusion, this project found that age and gender were important predictors of both healthcare utilization and healthcare coverage in a synthetic healthcare dataset. Older patients tended to have more encounters and higher coverage, while female patients showed higher utilization and higher average coverage than male patients. Race was not significant in the bivariate hypothesis tests, but some racial differences appeared in the regression models after adjustment for other predictors. Although the synthetic nature of the dataset limits real-world interpretation, the project demonstrates how sociodemographic inequalities in healthcare access and utilization can be investigated using statistical methods.
