---
title: "Sociodemographical factors predicting healthcare utilization and insurance coverage patterns"
author: "Lovisa Johansson & Roberta Merialdo"
format: html
---

::: {.callout-note}
This report was generated using synthetic data! It is only used for demonstration purposes and does not reflect real patients or healthcare providers. The contents have not been reviewed or validated by medical professionals.
:::

## Introduction

Understanding how sociodemographic factors influence healthcare utilization and insurance coverage is essential for identifying disparities in healthcare access and outcomes. Age, gender, and race are known to be associated with differences in health needs and healthcare-seeking behaviour. This project explores these relationships using a synthetic healthcare dataset, with the aim of identifying patterns that may reflect real-world inequalities in healthcare systems.

## Research Question
*"Do sociodemographic factors (age, gender, race) predict healthcare utilization and insurance coverage patterns?"*

## Data and Methods

### Dataset Description

This analysis uses synthetic healthcare data generated by [Synthea](https://github.com/synthetichealth/synthea), a synthetic patient population simulator. The dataset contains 17 interlinked CSV files covering patients, clinical encounters, conditions, medications, procedures, and insurance information. The primary datasets used in this analysis are:

- **patients** — demographic information including age, gender, race, and insurance coverage
- **encounters** — records of all clinical visits per patient, used to measure healthcare utilization
- **payer_transitions** — records of insurance coverage changes over time

### Variable Definitions

| Variable | Source | Description |
|---|---|---|
| `age` | patients | Age in years, derived from birthdate |
| `gender` | patients | Patient gender (M/F) |
| `race` | patients | Patient race (white, black, asian, etc.) |
| `n_encounters` | encounters | Number of clinical encounters per patient (outcome 1) |
| `healthcare_coverage` | patients | Total insurance coverage amount in USD (outcome 2) |

### Statistical Methods

Two outcomes were analysed:

1. **Healthcare utilization** (`n_encounters`) — a count variable modelled using Poisson regression
2. **Insurance coverage** (`healthcare_coverage`) — a continuous variable modelled using linear regression

Prior to regression, hypothesis tests were conducted for each predictor:

- **Age** — Spearman correlation (non-parametric, due to skewed distributions)
- **Gender** — Independent samples t-test
- **Race** — Kruskal-Wallis test (non-parametric, multiple groups)

## Data Preparation

```{r}
#| label: data-preparation
#| message: false

library(ggplot2)
library(knitr)
library(data.table)
targets::tar_load(dt_patients)
targets::tar_load(dt_encounters)
targets::tar_load(dt_payer_transitions)

# Calculate age from birthdate
dt_patients[, age := as.numeric(difftime(Sys.Date(), as.Date(birthdate), units = "days")) / 365.25]

# Count number of encounters per patient (healthcare utilization)
encounter_counts <- dt_encounters[, .(n_encounters = .N), by = patient]

# Join patients with encounter counts and coverage
analysis <- dt_patients[encounter_counts, on = .(id = patient), nomatch = 0]
analysis <- analysis[, .(id, age, gender, race, n_encounters, healthcare_coverage)]

# Remove deceased patients
analysis <- analysis[is.na(dt_patients[match(analysis$id, dt_patients$id), deathdate]) |
                     dt_patients[match(analysis$id, dt_patients$id), deathdate] == ""]
```

### Create Analysis Dataset

### Table of Missing Values in Key Variables

```{r}
#| label: missing-values
missing <- sapply(analysis, function(x) sum(is.na(x)))
kable(data.frame(Variable = names(missing), Missing = missing), row.names = FALSE)
```

## Exploratory Data Analysis

### Sample Characteristics

```{r}
#| label: sample-characteristics
summary_table <- analysis[, .(
  N = .N,
  Mean_Age = round(mean(age, na.rm = TRUE), 1),
  SD_Age = round(sd(age, na.rm = TRUE), 1),
  Mean_Encounters = round(mean(n_encounters, na.rm = TRUE), 1),
  Mean_Coverage = round(mean(healthcare_coverage, na.rm = TRUE), 0)
)]
kable(summary_table)
```

### Visualisations

@fig-age-distribution shows the age distribution of patients. @fig-encounters-gender and @fig-encounters-race display healthcare utilization by gender and race respectively. @fig-coverage-race shows insurance coverage by race, and @fig-age-encounters illustrates the relationship between age and number of encounters.

```{r}
#| label: fig-age-distribution
#| fig-cap: "Age distribution of patients"
ggplot(analysis, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(x = "Age (years)", y = "Count", title = "Age Distribution") +
  theme_minimal()
```

```{r}
#| label: fig-encounters-gender
#| fig-cap: "Number of encounters by gender"
ggplot(analysis, aes(x = gender, y = n_encounters, fill = gender)) +
  geom_boxplot() +
  labs(x = "Gender", y = "Number of Encounters", title = "Healthcare Utilization by Gender") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#| label: fig-encounters-race
#| fig-cap: "Number of encounters by race"
ggplot(analysis, aes(x = race, y = n_encounters, fill = race)) +
  geom_boxplot() +
  labs(x = "Race", y = "Number of Encounters", title = "Healthcare Utilization by Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

```{r}
#| label: fig-coverage-race
#| fig-cap: "Insurance coverage by race"
ggplot(analysis, aes(x = race, y = healthcare_coverage, fill = race)) +
  geom_boxplot() +
  labs(x = "Race", y = "Healthcare Coverage ($)", title = "Insurance Coverage by Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

```{r}
#| label: fig-age-encounters
#| fig-cap: "Relationship between age and number of encounters"
ggplot(analysis, aes(x = age, y = n_encounters)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red") +
  labs(x = "Age (years)", y = "Number of Encounters", title = "Age vs Healthcare Utilization") +
  theme_minimal()
```

## Hypothesis Testing

### Hypothesis 1: Age predicts healthcare utilization and coverage

```{r}
#| label: h1-age
# H1: Age predicts healthcare utilization and coverage
cor.test(analysis$age, analysis$n_encounters, method = "spearman")
cor.test(analysis$age, analysis$healthcare_coverage, method = "spearman")
```

### Hypothesis 2: Gender differences in utilization and coverage

```{r}
#| label: h2-gender
# H2: Gender differences in utilization and coverage
t.test(n_encounters ~ gender, data = analysis)
t.test(healthcare_coverage ~ gender, data = analysis)
```

### Hypothesis 3: Race differences in utilization and coverage

```{r}
#| label: h3-race
# H3: Race differences in utilization and coverage
kruskal.test(n_encounters ~ race, data = analysis)
kruskal.test(healthcare_coverage ~ race, data = analysis)
```

## Statistical Analysis

```{r}
#| label: regression

# Outcome 1: Healthcare utilization (count → Poisson regression)
model_utilization <- glm(n_encounters ~ age + gender + race,
                         data = analysis,
                         family = poisson())
summary(model_utilization)

# Outcome 2: Insurance coverage (continuous → linear regression)
model_coverage <- lm(healthcare_coverage ~ age + gender + race,
                     data = analysis)
summary(model_coverage)
```

## Discussion and Conclusion

### Key Findings

The analysis examined whether age, gender, and race predict healthcare utilization (number of encounters) and insurance coverage among synthetic patients. Results from the hypothesis tests and regression models reveal the extent to which these sociodemographic factors are associated with differences in healthcare access and use.

### Clinical Implications

If sociodemographic factors such as race and gender are significantly associated with healthcare utilization and coverage, this may indicate systemic disparities in healthcare access. Such findings — even in synthetic data — can motivate the development of more equitable healthcare policies and targeted interventions for underserved groups.

### Limitations

- The data is entirely synthetic and generated by Synthea. While it mimics real-world patterns, it does not reflect actual patients or healthcare providers, and conclusions cannot be generalised to real populations.
- The analysis does not account for confounding variables such as income, geographic location, or underlying health conditions, which may explain part of the observed associations.
- The Poisson regression assumes that the mean equals the variance in encounter counts. If overdispersion is present, a negative binomial model would be more appropriate.

### Conclusion

This project explored the relationship between sociodemographic factors and healthcare outcomes using a reproducible targets-based pipeline. The findings provide a foundation for further investigation into healthcare disparities and demonstrate the value of structured, reproducible data analysis workflows.
